---
title: "Exercise 2 — Gene expression to context-specific models"
subtitle: "Instructor answer key | expression evidence and context-model handoff"
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    css: ../course_toc.css
execute:
  echo: true
  warning: false
  message: false
---

## Implementation status

The worksheet loads both profiles, evaluates synthetic GPR examples through the shared evaluator, and runs the configured signed-flux iMAT-like MILP twice under the same Recon1 network, biomass objective and solver settings but **different** documented RPMI/Keibler recipe-proxy media. Context RDS files and provenance/QC CSVs are written only when both solver results and post-reconstruction checks pass. Gene evidence remains distinct from the optimized returned model. Numerical model outputs are generated by the worksheet rather than hard-coded here.

## 1. Profile identity and schema

Expected schemas are 1,905 Recon1 gene-product rows plus two identifier columns; each cancer file has 625 ACH values (627 columns total), while each breast file has one value/evidence column (3 total columns). The crosswalk contains one MCF7 match to `ACH_000019`, with status `matched` and a unique ID present in the DepMap matrix. This establishes the sample identifier cross-reference, not matched culture conditions or paired biological replicates. The GTEx breast input is one archived tissue-level profile, not a donor-matched comparator and not Keibler HMEC RNA.

**Question 1 answer.** The comparison is between two independent source contexts: MCF7, a cancer cell line, and one archived GTEx breast-tissue profile. The profiles are not matched biological replicates or samples from the same patient. Therefore the model comparison can illustrate different conditional network states, but it cannot establish a general cancer-versus-healthy biological difference.

**Run this check:**

```{r}
#| label: answer-check-profile-identity
setwd("..")
crosswalk_path <- "datasets_and_models/treated/cancer/transcriptomics/DepMap_Jain_NCI60_crosswalk.csv"
depmap_crosswalk <- read.csv(crosswalk_path, stringsAsFactors = FALSE)
depmap_crosswalk[depmap_crosswalk$depmap_sample_id == "ACH_000019", c(
  "jain_cell_line", "depmap_sample_id", "mapping_status"
)]
```

**Interpret the result:** the row should identify `MCF7` and `ACH_000019` with `matched` status. That confirms identity only; it does not confirm matched tissue, donor, culture conditions, or biological replicates.

## 2. Missingness, zeroes, and thresholds

Expected profile-level QC:

| Profile / measure | Finite/nonmissing | Zero | Missing | Evidence categories |
|---|---:|---:|---:|---|
| MCF7 expression (`ACH_000019`) | 1,694 | 161 | 211 | — |
| MCF7 gene evidence | 1,694 | — | 211 | `-1`: 426; `0`: 844; `1`: 424 |
| GTEx breast expression | 1,688 | 52 | 217 | — |
| GTEx breast gene evidence | 1,688 | — | 217 | `-1`: 422; `0`: 844; `1`: 422 |

The saved MCF7 thresholds are Q1 = **0.28 TPM** and Q3 = **39.4675 TPM**. They are sample-specific empirical cutoffs. The evidence categories summarize low/intermediate/high expression under the documented profile-level rule; blank means missing/unmapped, not low. The exporter uses the documented tie policy for equal quartiles. DepMap is already TPM; the breast column comes from the archived GTEx/tINIT `levels` profile and has a different processing history/scale.

Interpret the student QC tables separately: expression `expression_zero` counts exact zero-valued measurements; evidence `intermediate_zero` counts the discrete category `0`, not zero expression. `nonmissing` is finite expression in the expression table, and the sum of `-1`, `0`, and `+1` categories in the evidence table. For MCF7, 1,694 nonmissing expression values include 161 exact zeroes and 1,533 nonzero values; its 1,694 nonmissing evidence entries are 426 low + 844 intermediate + 424 high. In both profiles, `nonmissing + missing = 1,905` rows.

Replacing missing values by zero would conflate “not observed/mapped” with measured zero expression. Replacing them with `-1` would manufacture low-expression evidence. Either can change downstream GPR logic, especially AND rules. The two expression profiles' counts/units and cell/tissue sampling differ; numerical values should not be treated as a matched comparison.

**Run this check:**

```{r}
#| label: answer-check-profile-qc
setwd("..")
expression_summary <- function(x) {
  c(rows = length(x), nonmissing = sum(is.finite(x)),
    expression_zero = sum(x == 0, na.rm = TRUE), missing = sum(is.na(x)))
}
evidence_summary <- function(x) {
  c(rows = length(x), nonmissing = sum(!is.na(x)),
    low_minus1 = sum(x == -1, na.rm = TRUE),
    intermediate_zero = sum(x == 0, na.rm = TRUE),
    high_plus1 = sum(x == 1, na.rm = TRUE), missing = sum(is.na(x)))
}
cancer_expression <- read.csv("datasets_and_models/treated/cancer/transcriptomics/RECON1_DepMap_expression.csv", check.names = FALSE)
cancer_evidence <- read.csv("datasets_and_models/treated/cancer/transcriptomics/RECON1_DepMap_iMAT_evidence.csv", check.names = FALSE)
breast_expression <- read.csv("datasets_and_models/treated/healthy/transcriptomics/RECON1_GTEx_breast_expression.csv", check.names = FALSE)
breast_evidence <- read.csv("datasets_and_models/treated/healthy/transcriptomics/RECON1_GTEx_breast_iMAT_evidence.csv", check.names = FALSE)
profile_qc <- rbind(
  MCF7_expression = expression_summary(cancer_expression[["ACH_000019"]]),
  GTEx_breast_expression = expression_summary(breast_expression[["GTEx_breast_tINIT_level"]])
)
evidence_qc <- rbind(
  MCF7_gene_evidence = evidence_summary(cancer_evidence[["ACH_000019"]]),
  GTEx_breast_gene_evidence = evidence_summary(breast_evidence[["GTEx_breast"]])
)
threshold_row <- read.csv("datasets_and_models/treated/cancer/transcriptomics/RECON1_iMAT_evidence_thresholds.csv", stringsAsFactors = FALSE)
profile_qc
evidence_qc
threshold_row[threshold_row$sample_id == "ACH_000019", c("low_threshold", "high_threshold")]
```

**Interpret the result:** `rows` is the aligned model-entry count, not a count of measured biological genes. For expression, `expression_zero` means an observed exact zero. For evidence, `intermediate_zero` means category `0`. Missing values remain unknown and are not silently converted into low expression.

## 3. Synthetic GPR truth examples

Expected shared evaluator results:

Each row is one hypothetical reaction. The GPR is evaluated using the listed gene-product evidence, and `Result` is the single reaction-evidence value returned for that GPR. The first three rows correspond directly to the three hand-calculation rows in the worksheet; the last two are additional missing-evidence policy checks.

| GPR | Evidence | Result | Reason |
|---|---|---:|---|
| `G_A and G_B` | `1`, `0` | `0` | Known AND combines by minimum. |
| `G_C or G_D` | `-1`, `1` | `1` | Known OR combines by maximum. |
| `(G_E and G_F) or G_G` | `1`, `NA`, `-1` | `NA` | The AND is unresolved; OR with `-1` cannot resolve that uncertainty. |
| `G_low and G_F` | `-1`, `NA` | `-1` | A known low AND partner is decisive despite missing evidence. |
| `G_C or G_F` | `1`, `NA` | `1` | A known high OR partner is decisive despite missing evidence. |
| Empty GPR | — | `NA` | No gene rule means unevaluated/unknown, not zero. |

The shared contract suite exercises the defined missing-evidence policy, malformed-input rejection, and a synthetic signed-flux MILP. Shared API `reaction_evidence_from_gpr(gpr, gene_evidence)` evaluates a restricted logical expression without `eval(parse())`; unknown or absent evidence remains unknown except for decisive AND-low / OR-high cases.

**Run this check:**

```{r}
#| label: answer-check-gpr-policy
setwd("..")
source("R/model_io.R")
fixtures <- list(
  and_intermediate = list(gpr = "G_A and G_B", evidence = c(G_A = 1, G_B = 0)),
  or_high = list(gpr = "G_C or G_D", evidence = c(G_C = -1, G_D = 1)),
  nested_unknown = list(gpr = "(G_E and G_F) or G_G", evidence = c(G_E = 1, G_F = NA, G_G = -1)),
  and_decisive_low = list(gpr = "G_low and G_F", evidence = c(G_low = -1, G_F = NA)),
  or_decisive_high = list(gpr = "G_C or G_F", evidence = c(G_C = 1, G_F = NA))
)
fixture_results <- vapply(fixtures, function(z) {
  reaction_evidence_from_gpr(z$gpr, z$evidence)
}, numeric(1))
data.frame(gpr = vapply(fixtures, `[[`, character(1), "gpr"),
           reaction_evidence = unname(fixture_results),
           row.names = names(fixture_results))
```

**Expected result:** the five reaction-level results are `0`, `1`, `NA`, `-1`, and `1` in table order.

**Interpretation:** AND combines known values by minimum and OR by maximum. A known `-1` makes an AND result `-1`; a known `+1` makes an OR result `1`. The nested example remains unknown because neither branch resolves the missing evidence. An empty GPR returns `NA`, meaning unevaluated/unknown, not intermediate evidence `0`.

## 4. Native exchange contexts and provenance

The primary comparison uses MCF7 expression with a Jain NCI-60 RPMI recipe proxy and GTEx breast expression with the 42-row Exercise 1 Keibler-derived bounds. Jain/Keibler measured exchange outcomes and direction summaries are not used to construct either medium. The RPMI proxy uses the standard RPMI-1640 reference concentrations (11.1 mM glucose, amino acids, vitamins, salts) with Jain's reported 2 mM glutamine, all capped with the same inventory factor as the Exercise 1 Keibler medium (same assumed volume and biomass-time). The Keibler medium uses Keibler's 10 mM glucose and 4 mM glutamine plus standard DMEM concentrations. Exact catalog formulations, MCF7-specific biomass-time, and Jain flask-volume/exposure conversion are unresolved; FBS is not represented. This is an explicit pedagogical scenario, not a quantitative reconstruction of the Jain assay. O2, CO2, H2O, and H+ share a 10 mmol/gDW/h uptake setting, and listed exchanges other than glucose and glutamine may secrete.

Both contexts use the same hands-on model hash, biomass objective, and iMAT settings, but different scenario-specific exchange constraints. `config/imat_settings.yml` pins the open-source HiGHS R interface (package 1.14.0.2, native engine 1.14.0), epsilon 0.1 mmol/gDW/h, biomass fraction 0.9, flux cap 1000, 300-second limit, two threads, equal positive/low weights, neutral intermediate/unknown evidence, and signed reversible-flux indicators. This is the configured iMAT-like implementation for this teaching exercise, not a universal parameter default.

Verified run results from the persisted QC (generated, not hard-coded):

| Context | MILP status | iMAT evidence objective | Biomass at selected MILP solution | Post-reconstruction FBA | Maximum $|S v|$ |
|---|---|---:|---:|---|---:|
| MCF7 (`ACH_000019`) + Jain constraints | generated from persisted QC | generated from persisted QC | generated from persisted QC | generated from persisted QC | generated from persisted QC |
| GTEx breast + Keibler bounds | generated from persisted QC | generated from persisted QC | generated from persisted QC | generated from persisted QC | generated from persisted QC |

These are solver/model outputs conditional on the declared expression and exchange scenarios, not measured growth rates. The two contexts are independent source representations, not matched tumor/healthy samples. Both MILP biomass values must meet the configured 90% of their own scenario-specific WT optimum; the post-reconstruction FBA optima may exceed that minimum.

**Run this inspection:**

```r
qc <- read.csv("outputs/reconstruction_qc.csv", stringsAsFactors = FALSE)
qc[, c("scenario_id", "medium_id", "objective_id", "objective_sense",
       "solver", "milp_status", "incumbent_feasible", "imat_objective",
       "incumbent_biomass_value", "reconstructed_fba_status",
       "reconstructed_fba_objective_value", "max_abs_sv")]
```

**Expected result:** two rows have `optimal` MILP status, feasible incumbents, `optimal` reconstructed FBA status, positive biomass, and `max_abs_sv <= 1e-7`. The objective, objective direction, solver, epsilon, and growth fraction are shared through the settings file. The `medium_id` values differ by design: `Jain_NCI60_RPMI_recipe_proxy_v2` identifies the assumption-based RPMI scenario and `Keibler_184A1_EV_EGF5_growth_adjusted_v2` identifies the Keibler scenario.

**Check the effective exchange bounds:**

```r
qc <- read.csv("outputs/reconstruction_qc.csv", stringsAsFactors = FALSE)
bounds <- read.csv("outputs/effective_exchange_bounds.csv", stringsAsFactors = FALSE)
by_context <- split(bounds, bounds$context)
lapply(by_context, function(x) {
  x[x$reaction_id %in% c("R_EX_glc__D_e", "R_EX_gln__L_e", "R_EX_lac__L_e"), ]
})
```

**Interpretation:** the two scenarios differ wherever the RPMI and DMEM reference concentrations differ (glucose, glutamine, amino acids such as asparagine/aspartate/glutamate/proline that are present only in RPMI, phosphate, iron, vitamins); O2, CO2, H2O, and H+ settings are equal. In both scenarios glucose is taken up to its cap in the selected optimum. MCF7 uses recipe-proxy bounds, not a Jain-informed envelope. The table is the capped medium scenario before the iMAT inactivity limits; the integrated RDS model additionally limits MILP-inactive low-evidence reactions to ±epsilon (active reactions are not forced). This is a combined expression-plus-medium-assumption comparison.

**Check the stored models:**

```r
mcf7 <- readRDS("outputs/context_MCF7_ACH_000019_Jain.rds")
healthy <- readRDS("outputs/context_GTEx_breast_Keibler.rds")
c(mcf7$provenance$scenario_id, healthy$provenance$scenario_id)
c(mcf7$provenance$medium_id, healthy$provenance$medium_id)
file.exists(c("outputs/context_MCF7_ACH_000019_Jain.xml",
              "outputs/context_GTEx_breast_Keibler.xml"))
```

The RDS files are the full-fidelity artifacts. They retain `scenario_model`/`scenario_exchange_bounds` separately from the iMAT-integrated `model`. The XML files contain the network, objective, and capped scenario exchange-medium bounds; they do not contain the iMAT inactivity limits or all R-specific QC/provenance fields.

### Reconstructed reaction counts — independent verification

**Run this** from a clean session. Count the actual integrated RDS bounds, not reactions absent from the XML (the XML retains the original network). Check the manifest-pinned full model and status-valid fresh WT FBA solutions; the full source row has no matching fixed medium, so its selected-flux entry is missing.

```{r}
#| label: answer-check-reaction-counts
root <- c(".", "..")
root <- root[file.exists(file.path(root, "R/model_io.R"))][1]
if (is.na(root)) stop("Open from exercises/ or the Exercise 2 folder.")
setwd(root)
source("R/model_io.R")
original <- load_model("datasets_and_models/models/RECON1_biomass.xml", "config/model_manifest.csv")
settings <- read_imat_settings("config/imat_settings.yml")
saved <- list(MCF7 = readRDS("Task2_expression_integration/outputs/context_MCF7_ACH_000019_Jain.rds"),
  GTEx = readRDS("Task2_expression_integration/outputs/context_GTEx_breast_Keibler.rds"))
count_one <- function(model, label, fba = NULL) {
  stopifnot(identical(model$reaction_ids, original$reaction_ids),
    identical(model$source_sha256, original$source_sha256))
  low <- model$lb >= -settings$epsilon - 1e-8 & model$ub <= settings$epsilon + 1e-8
  if (!is.null(fba)) stopifnot(fba$status == "optimal", fba$objective_value > 0,
    fba$max_abs_sv <= fba$tolerance)
  data.frame(model = label, total_reaction_ids = length(model$reaction_ids),
    bounds_closed_at_zero = sum(model$lb == 0 & model$ub == 0),
    limited_to_plus_minus_epsilon = sum(low),
    capacity_above_epsilon = sum(!low),
    selected_fba_nonzero_above_1e_minus_7 = if (is.null(fba)) NA_integer_ else
      sum(abs(fba$fluxes[model$reaction_ids]) > 1e-7))
}
counts <- do.call(rbind, list(
  count_one(original, "Full Recon1 source (no fixed medium)"),
  count_one(saved$MCF7$model, "MCF7 / RPMI proxy", solve_fba(saved$MCF7$model)),
  count_one(saved$GTEx$model, "GTEx breast / Keibler proxy", solve_fba(saved$GTEx$model))))
stopifnot(all(counts$total_reaction_ids == 3742L),
  identical(as.integer(counts$limited_to_plus_minus_epsilon[2:3]),
    c(saved$MCF7$evidence_qc$negative_inactive, saved$GTEx$evidence_qc$negative_inactive)))
saved_counts <- read.csv("Task2_expression_integration/outputs/reaction_count_comparison.csv",
  stringsAsFactors = FALSE)
stopifnot(isTRUE(all.equal(counts, saved_counts, check.attributes = FALSE)))
counts
```

**Expected result:** 3,742 reaction IDs and zero bounds closed at zero in each row. The full source has all 3,742 reactions capable of $|v|>0.1$ under its raw bounds, with no comparable selected FBA point. MCF7 has 358 reactions limited to ±0.1, 3,384 with capacity above 0.1, and 646 nonzero in one selected FBA solution ($|v|>10^{-7}$). GTEx/Keibler has 297 limited, 3,445 with capacity above 0.1, and 658 selected nonzero. Check the saved `outputs/reaction_count_comparison.csv` against this independently calculated table.

**Interpretation:** "Limited" is not "knocked out" or "absent": iMAT preserves residual flux capacity. A single FBA optimum does not establish that all remaining reactions are used, or that MCF7 is universally more vulnerable. The different proxy media confound biological comparisons.

Use `outputs/reconstruction_qc.csv` for actual solver and evidence-QC results, `outputs/effective_exchange_bounds.csv` for capped pre-iMAT scenario-medium bounds, `outputs/context_MCF7_ACH_000019_Jain.rds` / `outputs/context_GTEx_breast_Keibler.rds` for full-fidelity reusable model states, and the matching `.xml` files for interoperable SBML network exports. The RDS files include source, evidence, threshold, scenario, medium hashes/assumptions, solver status, selected fluxes, and provenance; XML contains the network, objective, and scenario-medium bounds but not the full R-specific MILP/QC metadata. A constraint-verified time-limit incumbent may be saved only with status `time_limit`, with incumbent biomass distinguished from a proven optimum; missing/infeasible incumbents or failed post-reconstruction QC stop the worksheet.

The current primary comparison combines expression and extracellular exchange-context effects. Spent-medium observations and extracellular rates do not by themselves establish a complete intracellular flux vector.

## 5. Answers to interpretation questions

1. **Expression to model chain.** Run `profile_qc` and `evidence_qc`; inspect `fixture_results`; then inspect `qc[, c("scenario_id", "milp_status", "imat_objective", "reconstructed_fba_status", "max_abs_sv")]`. Expression values become gene-product evidence, GPR rules become reaction evidence, and the MILP uses those categories together with stoichiometry, exchange bounds, and the biomass floor. The solver status and post-reconstruction FBA residual support computational validity, not experimental validation.
2. **Repeated symbols.** Run `table(cancer_expression$gene_symbol)` and inspect the aligned `recon1_gene_id` values. Multiple model gene-product IDs can share a symbol and therefore inherit related source evidence, but they can participate in different GPRs and reactions. Shared evidence does not prove shared enzyme complexes, reaction rates, or identical fluxes.
3. **Why medium context matters.** Run `unique(qc$medium_id)` and compare the two rows in `effective_exchange_bounds.csv`. The primary comparison changes both expression evidence and medium assumptions: MCF7 uses the RPMI recipe proxy, while GTEx breast uses the Exercise 1 Keibler-derived proxy. Therefore it is a combined-context comparison, not an expression-only experiment.

**Reconstruction check answer.** The model, biomass objective, objective direction, iMAT settings, solver, epsilon, and growth fraction are shared. The scenario exchange bounds differ wherever the two reference formulations differ: MCF7 uses the assumption-based RPMI recipe proxy, while GTEx breast uses the Exercise 1 Keibler-derived bounds. The iMAT objective rewards high-evidence reactions for signed flux above epsilon and low-evidence reactions for flux within the inactive threshold; intermediate and unknown evidence are neutral. An optimal MILP only demonstrates that the mathematical network has a feasible flux assignment satisfying the declared stoichiometry, medium bounds, biomass floor, and evidence objective. It does not make MCF7 and GTEx matched biological samples, and its biomass flux is a model output rather than a measured growth rate. Jain/Keibler exchange outcomes remain held out for subsequent assessment.
